A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-05 rigid_bodies.tex
Update of splitting equations

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^t}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

@@

\usepackage{booktabs}
\usepackage{multirow}

@@

@@

As done in the Hamiltonian case, the evolution of an extended phase space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the Liouville operator can be written as
\[
i\!L = i\!L_\text{NHC} + i\!L_\text{NVE},
\]
where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
\begin{equation}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) \\
&+ \sum_{j=1}^M \left[ \alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi(\alpha_{j+1} t) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$. Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$. Considering the analytical solution above, the NHC operator is factorized as
\begin{widetext}
\begin{equation}
\begin{split}
e^{i\!L_\text{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&  \times \prod_{j=1}^{M} \exp\left(\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)  \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt p}_i \diff{}{\vt p_i}\right) \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)  \\
& \times \prod_{j=1}^{M} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
\end{split}
\end{equation}
\end{widetext}
The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
\begin{equation}
\vt \pi_i \leftarrow \vt \pi_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
\end{equation}
\begin{equation}
\vt p_i \leftarrow \vt p_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
\end{equation}
\begin{equation}
\eta_j  \leftarrow \eta_j + \frac{p_{\eta_j}}{ Q_j} \frac{\Delta t}{2} 
\end{equation}

@@

\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{i L \Delta t} =
e^{i L_{NHC} \frac{\Delta t}{2}} e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t}  e^{i L_a \frac{\Delta t}{2}} e^{i L_{NHC} \frac{\Delta t}{2}}.
\end{equation}

@@

\begin{equation}
\label{eq:pi}
\begin{split}
{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t,
\end{split}
\end{equation}
respectively, with $\phi(x)$ given by Eq.~\ref{eq:phi}. Therefore, one of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered. For that, let as define the modified propagators $ e^{i L^\ast_a} $ and $ e^{i L^\ast_{NHC}} $ as follows:
%\begin{equation}
%\begin{split}
%&i L^\ast_a = i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1} \vt p_i^T \diff{}{\vt p_i} - %\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i^T  \diff{}{\vt \pi_i} =\\
%&\sum_{i=1}^{N} \left(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} + %\sum_{i=1}^{N} \left(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T\right)  \diff{}{\vt %\pi_i}
%\end{split}
%\end{equation}
%\begin{equation}
%\begin{split}
%i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} %\frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
%\end{split}
%\end{equation}
\begin{equation}
\begin{split}
e^{iL^\ast_a \frac{\Delta t}{2}} = &\prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left( \vt F_i^T - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} \right] \\
&\times \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left(2 \vt \tau_i^T \mt C_i^T -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T \right)  \diff{}{\vt \pi_i}   \right]
\end{split}
\end{equation}

\begin{equation}
\begin{split}
e^{iL^\ast_{NHC} \frac{\Delta t}{2}} = &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&\times\prod_{j=1}^{M} exp\left(-\frac{\Delta t}{2}  \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)   \\
&\times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
\end{split}
\end{equation}
Finally, the complete propagator scheme is
\begin{equation}
\label{eq:trotter_splitting_*}
\begin{split}
e^{i L \Delta t} = &e^{i L^\ast_{NHC} \frac{\Delta t}{2}} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L^\ast_{NHC} \frac{\Delta t}{2}}.
\end{split}
\end{equation}

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}

\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\usepackage{booktabs}
\usepackage{multirow}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^t}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

@@

@@

\[
\begin{split}
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
\end{split}
\]

The same actions, but simultaneously for all rigid bodies present in the system.

@@

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the Liouville operator can be written as
\[
i\!L = i\!L_\text{NHC} + i\!L_\text{NVE},
\]
where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
\begin{equation}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) \\
&+ \sum_{j=1}^M \left[ \alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$. Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$. Considering the analytical solution above, the NHC operator is factorized as
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left(G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&\times \prod_{j=1}^M \exp\left(\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \alpha_1 \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)\right] \\
&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
\end{split}
\end{equation}

The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
\begin{equation}
\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
\end{equation}
\begin{equation}
\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
\end{equation}
\begin{equation}
\eta_j = \eta_j^0 + \alpha_j^0 \frac{\Delta t}{2} 
\end{equation}

@@

\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
\end{equation}

@@

\begin{equation}
\label{eq:pi}
\begin{split}
{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t,
\end{split}
\end{equation}
respectively, with $\phi(x)$ given by Eq.~\ref{eq:phi}. Therefore, one of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered. For that, let as define the modified propagators $ e^{i L^\ast_a} $ and $ e^{i L^\ast_{NHC}} $ as follows:
%\begin{equation}
%\begin{split}
%&i L^\ast_a = i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1} \vt p_i^T \diff{}{\vt p_i} - %\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i^T  \diff{}{\vt \pi_i} =\\
%&\sum_{i=1}^{N} \left(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} + %\sum_{i=1}^{N} \left(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T\right)  \diff{}{\vt %\pi_i}
%\end{split}
%\end{equation}
%\begin{equation}
%\begin{split}
%i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} %\frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
%\end{split}
%\end{equation}
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_b^\ast} &= \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \right] \\
&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left(2 \tr{\vt \tau}_i \tr{\mt C}_i -  \alpha_1 \tr{\vt \pi}_i \right)  \diff{}{\vt \pi_i}   \right]
\end{split}
\end{equation}

\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&\times \prod_{j=1}^M \exp\left(-\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
\end{split}
\end{equation}
Finally, the complete propagator scheme is
\begin{equation}
\label{eq:trotter_splitting_*}
\begin{split}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast}.
\end{split}
\end{equation}
